% !TEX TS-program = arara
% arara: xelatex: { synctex: on, options: '-halt-on-error' } 
% arara: biber
% arara: makeglossaries
% arara: makeindex
% arara: xelatex: { synctex: on, options: '-halt-on-error'  } 
% arara: xelatex: { synctex: on, options: '-halt-on-error'   } 
% arara: clean: { files: [ mthcmptng-book.aux, mthcmptng-book.bbl] }
% arara: clean: { files: [ mthcmptng-book.bcf, mthcmptng-book.blg] }
% arara: clean: { files: [ mthcmptng-book.glg, mthcmptng-book.glo] }
% arara: clean: { files: [ mthcmptng-book.gls, mthcmptng-book.gls] }
% arara: clean: { files: [ mthcmptng-book.idx, mthcmptng-book.ilg] }
% arara: clean: { files: [ mthcmptng-book.ind, mthcmptng-book.loe] }
% arara: clean: { files: [ mthcmptng-book.lof, mthcmptng-book.log ] }
% arara: clean: { files: [ mthcmptng-book.log, mthcmptng-book.lol ] }
% arara: clean: { files: [ mthcmptng-book.out ] }
% arara: clean: { files: [ mthcmptng-book.run.xml] }
% arara: clean: { files: [ mthcmptng-book.toc, mthcmptng-book.xdy] }
% arara: clean: { files: [ mthcmptng-book.synctex.gz] }
%-------------------------------------------------------------------------------
\documentclass[10pt,openany]{book}
\input{head}
\input{landscape-2col-header}
\input{clj-listings}
\input{glossary}
\addbibresource{cactus.bib}
\addbibresource{clojure.bib}
\addbibresource{halmos.bib}
\addbibresource{ieeestd.bib}
\addbibresource{tex.bib}
\addbibresource{fparith.bib}
%-------------------------------------------------------------------------------
\title{Math Computing (working title)}
\author{\textsc{John Alan McDonald}}
%\email{mcdonald.john.alan at gmail.com}
\date{draft of \today}
%-------------------------------------------------------------------------------
\begin{document}

\maketitle

\frontmatter

\begingroup
\let\onecolumn\twocolumn
\sffamily
\tableofcontents
\rmfamily
\endgroup

\mainmatter

\part{Aspirations and Inspirations}
\chapter{Good writing}

\index{Good writing|see{Set}}
Primary goal of writing, whether in a natural language (eg
English), mathematics, or a programming language, is communicating
an idea to the reader.
This is (perhaps) obvious for natural and mathematical language,
but not univerally accepted for programming languages.

\cite{Halmos1970HowToWrite}

('Literate' programming confuses stream of consciousness writing
with communication with other people.)

Express the idea completely enough, without excess.
and specify no more than
necessary --- the he/she problem.
\index{Type problems@\textsl{Sam}!he/she@\textbf{he/she}}
Make it clear what you aren't making clear.

\part{Foundations}

Most readers will be familiar with most of what's here, 
but it's worth at least skimming, 
because the point of view is different.

\chapter{Mathematics}
\section{Ambiguity}
\section{Abstraction, instantiation, and representation}

pragmatic, semi-axiomatic definition of things.

\chapter{A model for computation}		

Transformation of data structures.

\section{Data Structures}

A \textit{data structure} is a collection of \textit{places}
that hold values. 
Using Clojure/Java for specificity.
A value might be primitive
(\texttt{int}, \texttt{double}, \ldots)
or a references to an 'independent' data structure.

\begin{example}[Java array]
Places correspond to \texttt{int}s from 0 to length minus 1.
Values restricted to instances of array's type.
\end{example}  

\begin{example}[Instance of Java class]
Places correspond to the class's fields.
Values restricted to each field's type.
\end{example}  

\begin{example}[EDN data]
Nested sequences and hashmaps. Similar to JSON.

Places correspond to the class's fields.
Values restricted to each field's type.
\end{example}  

\subsection{Mutability}
\subsection{Random access}

Means different things. 
Constant time to any place.
Absolute origin, rather than relative references.
Places known/easily enumerable, rather than requring traversal
to find out what places there are.
No general guarantee that you can visit everything, once.

\subsection{An unsolved problem}

Where's the boundary of a thing? 
Is a reference part of a path, or
an encapsulated value?
No language I know does this well.

Deep copy vs shallow copy.


\section{Code and data}

Data is just bits. 
Meaning determined by the code that manipulates it. 

Code and meaning always changing.

Correctness difficulty increases
with 'distance' between code and data. 

Multiple implementations of meaning really bad.
Data standards (eg IEEE 754 \cite{Higham2002ASNA, IEEE:1985:AIS,
P754:2008:ISF, Muller-et-al-2010}) major undertaking even for
small simple unchanging data semantics.

Services bad; shared libraries good.
Crossing programming language boundaries bad; 
monolingual environments \cite{Heering:1985:TMP:3318.3321} good.

\section{Types and prototypes}

The ``he or she'' problem: forced to specify irrelevant details.




\section{Clojure and Java}

\chapter{Comparisons}

\part{Spaces and functions between them}

I'm assuming that you are generally familiar with things like:
\begin{itemize}
  \item what a \gls{Set} is, at least in general terms.
  \item the usual number systems: 
the natural numbers, the integers, the rationals, and the reals, 
and the arithmetic operations ($+$, $-$, $*$, $/$) on them.
\item the difference between the
countable infinities of the natural, integral, and rational numbers,
and the difference from the uncountable infinity of the real numbers.
\end{itemize} 

If not, it would be good to review some basic source, like Wikipedia.

In the rest of this book, I'm going to use the word 'space' to refer to
mathematical structures, like the usual number systems, vector and affine
spaces, groups, fields, \ldots, that consist of one, or two, or a few, sets,
plus operations on the elements of those sets.

The first concrete examples of this will be the usual number systems 
in~\autoref{ch:Numbers}.

\input{sets}
\input{functions}
\input{numbers}

\chapter{Simplexes}
\section{mathematics}
\subsection{Orientation}
\section{computation}
\section{examples}

\chapter{Linear (vector) spaces and functions}
\section{mathematics}
\cite{Halmos1960Finite}
\section{computation}
\section{examples}

\chapter{Affine (flat) spaces and functions}
\section{mathematics}
\section{computation}
\section{examples}

\chapter{Projective spaces and functions}
\section{mathematics}
\section{computation}
\section{examples}

\chapter{Barycentric (convex) spaces and functions}
\section{mathematics}
\section{computation}
\section{examples}

\chapter{Spherical spaces and functions}
\section{mathematics}
\section{computation}
\section{examples}


%\chapter{Notation and general results}
%\input{general}

%\part{Smoothness}
%\cleardoublepage

%\chapter{Polylines}
%\input{curves}
%\cleardoublepage

%\chapter{Triangular Meshes}
%\input{faces}
%\input{edges}
%\input{vertices}
%\cleardoublepage

%\chapter{Subdivision Surfaces}
%\input{subdivision}

%\part{Fitting}
%\cleardoublepage

%\chapter{Data Fitting}
%\input{data-fitting}
%\cleardoublepage

%\chapter{Image Fitting}
%\input{image-fitting}
%\cleardoublepage

%\chapter{Registration}
%\input{registration}
%\cleardoublepage

%\chapter{Averaging}
%\input{averaging}

\glsaddallunused
%-------------------------------------------------------------------------------
\appendix
\part{Appendices}
\input{typesetting}
\input{annoyances}
%-------------------------------------------------------------------------------
\backmatter


\part{Backmatter}
\input{tail}
%-------------------------------------------------------------------------------
\end{document}
